---
title: "1 skyriaus uzdaviniai"
output: html_document
date: "2023-09-10"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(fBasics)
library(markdown)
library(dplyr)
```

# 1
### a
```{r}
data_1.1 <- read.table("d-3stock.txt",header=F) 
names(data_1.1) <- c("date", "axp", "cat", "sbux")
Data_1.1 <- data_1.1 %>% 
  mutate(date = paste0(substr(date,1,4),"-",substr(date,5,6),"-",substr(date,7,8)))
sdata_1 <- Data_1.1[,2:4]*100  # simple returns in percentages
head(sdata_1)
basicStats(sdata_1) #sample mean, st dev, skewness, excess kurtosis, min, max
```

### b
```{r}
lData_1.1 <- log(Data_1.1[,2:4]+1) #logreturns
head(lData_1.1)
```

### c
```{r}
lData_1.1_perc <- lData_1.1*100 #log-returns in percentages
basicStats(lData_1.1_perc)
```

### d
```{r}
t.test(lData_1.1[,1]) #axp
t.test(lData_1.1[,2]) #cat
t.test(lData_1.1[,3]) #sbux
```
For 3 selected stock returns, p-value > 0.05, therefore with 5% significance level test fails to reject hypothesis, that expected means are equal to 0.

# 2
### a
```{r}
data_1.2 <- read.table("m-ibm3dx7503.txt",header=F) 
names(data_1.2) <- c("date", "ibm", "vw", "ew","sp")
Data_1.2 <- data_1.2 %>% 
  mutate(date = paste0(substr(date,1,4),"-",substr(date,5,6),"-",substr(date,7,8)))
sdata_2 <- Data_1.2[,2:5]*100  # simple returns in percentages
head(sdata_2)
basicStats(sdata_2) #sample mean, st dev, skewness, excess kurtosis, min, max
```

Toliau tas pats kaip 1 uždavinyje.

# 3
### a
Average annual return = sample mean * 12 = (sum of log returns)/(number of years)
```{r}
lsp <- log(Data_1.2[,5]+1)
annual_return_sp <- sum(lsp)/(2003-1975+1) #average annual log return of S&P
annual_return_sp
```

### b
```{r}
exp(1*annual_return_sp*(2003-1975+1))
```
Value of the investment would be 16.22 USD.

# 4
### a
```{r}
data_1.4 <- Data_1.1[,1:2]
ldata_1.4 <- log(data_1.4[,2]+1)
s1 <- skewness(ldata_1.4)
t1 <- s1/sqrt(6/length(ldata_1.4)) #Compute test statistic
t1
pv1 <- 2*(1-pnorm(abs(t1))) # Compute p-value.
pv1
```

`H0: S(r)=0, H1: S(r) != 0, |t-ratio| = 1.26 < 1.96`, also `p-value = 0.2 > 0.05`, therefore test fails to reject null hyphothesis, that skewness is not different from 0. 

### b
```{r}
k1 <- kurtosis(ldata_1.4)
t2 <- (k1-3)/sqrt(24/length(ldata_1.4)) #Compute test statistic
t2
pv2 <- 2*(1-pnorm(abs(t2))) # Compute p-value.
pv2
```
`H0: K(r)=0, H1: K(r) != 0, |t-ratio| = 4.98 > 1.96`, also `p-value < 0.05`, therefore test rejects null hyphothesis, indicating that kurtosis is different from 0. 

# 5
### a
