---
title: "Tsay_Uni_3_skyrius"
output: html_document
date: "2023-09-19"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(fBasics)
library(markdown)
library(dplyr)
library(rugarch)
library(fGarch)
library(tseries)
```

# 1,2,3,4
Sasiuvinyje

# 7
# a
```{r}
data_3.7 <- read.table("m-3m4608.txt",header=T) 
# names(data_3.7) <- c("date", "3M")
Data_3.7 <- data_3.7 %>% 
  mutate(date = paste0(substr(date,1,4),"-",substr(date,5,6),"-",substr(date,7,8)))
log_data_3.7 <- log(Data_3.7[,2]+1)  # log returns

#Checking for serial correlations (r)

lag_order <- 6
ljung_box_test <- Box.test(log_data_3.7, lag = lag_order, type = "Ljung-Box")
ljung_box_test

lag_order <- 12
ljung_box_test <- Box.test(log_data_3.7, lag = lag_order, type = "Ljung-Box")
ljung_box_test

# Checking for ARCH effect (r^2)
lag_order <- 6
ljung_box_test <- Box.test(log_data_3.7^2, lag = lag_order, type = "Ljung-Box")
ljung_box_test

lag_order <- 12
ljung_box_test <- Box.test(log_data_3.7^2, lag = lag_order, type = "Ljung-Box")
ljung_box_test
```
The Ljung-Box statistics of the squared log residuals shows that both Q(6) and Q(12) have pvalue < 0.05. Thus, there is signiﬁcant ARCH eﬀect in the log return series.

Based on the 5% signiﬁcance level, the log return series has serial correlations and has strong ARCH eﬀects.


### b
```{r}
plot(pacf(log_data_3.7^2))
```
From the plot it seems like we should specify ARCH(2)

```{r}
m1 <- garchFit(~arma(0,0)+garch(2,0),data=log_data_3.7, trace = FALSE)
summary(m1)

#Perform acf of model residuals to check model adequacy
residuals <- residuals(garchFit(~arma(0,0)+garch(2,0),data=log_data_3.7, trace = FALSE))
acf(residuals)
# pacf(residuals)
```
Acf plot still shows autocorrelation in lag 1 and 2, the model should be adjusted. Considering 10% significance level, the specified model $r_{t} = 0.012 + a_{t}$, $a_{t} = \sigma_{t} \epsilon_{t}$, where $\epsilon_{t} ~ N(0.1)$
$$ \sigma_{t}^2 =  aprasyti su abiem lagais $$

### c
```{r}
# Create GARCH model for in-sample period
data_3.7_insample <- log_data_3.7[1:750]
m2 <- garchFit(~arma(0,0)+garch(2,0),data=data_3.7_insample,trace=FALSE)
summary(m2)
predict(m2,5)
```

### d
We fit ARCH(2)-M model which is $r_{t} = \mu + c \sigma_{t}^2 + a_{t}$, $a_{t} ~ N(0.1)$,
$\sigma_{t}^2 = \alpha_{0} + \alpha_{1} a_{t-1}^2 + \alpha_{2} a_{t-2}$. c is considered risk premium.


```{r}
#NEVEIKIA
m3 <- garchOxFit(formula.mean=~arma(0,0),formula.var=~garch(0,2),series=data_3.7_insample,arch.in.mean=T)
```
p-value >0.05, therefore under 5% confidence risk premium is not different from 0.

### e
```{r}
# will be performed using rugarch, since fGarch doesn't have this option. 
spec <- ugarchspec(variance.model = list(model = "eGARCH", garchOrder = c(1, 1)),
                   mean.model = list(armaOrder = c(0, 0), include.mean = TRUE),
                   distribution.model = "std")
fit <- ugarchfit(spec, data = data_3.7_insample)
fit
ugarchforecast(fit, n.ahead = 5)
```

Leverege effect and asymmetry parameters are statistically insignificant, therefore there doesn't exist asymettry and EGARCH model is not correct model for the selected data.

Performed 5 step ahead forecast, but it shouldn't be trusted since model is not adequate.


# 8
### a
First, we check log-returns to evaluate which lags to use
```{r}
data_3.8 <- read.table("m-gmsp5008.txt",header=T) 
Data_3.8 <- data_3.8 %>% 
  mutate(date = paste0(substr(date,1,4),"-",substr(date,5,6),"-",substr(date,7,8)))
log_data_3.8 <- log(Data_3.8[,2:3]+1) # log returns

lag_order <- 6
ljung_box_test <- Box.test(log_data_3.8[,1], lag = 6, type = "Ljung-Box")
ljung_box_test

pacf(log_data_3.8[,1])

```
There exists serial dependence in log returns. Checking PACF shows that there are correlations in a few higher lags (14,25) and slight correlation in first two lags. Decide to proceed with mean model ARMA(0,0) and move to GARCH.

```{r}
spec_3.8.1 <- ugarchspec(variance.model = list(model = "sGARCH", garchOrder = c(1,1)),
                         mean.model = list(armaOrder = c(0,0), include.mean = TRUE),
                         distribution.model = "norm")
fit_3.8.1 <- ugarchfit(spec_3.8.1, log_data_3.8[,1])
fit_3.8.1
```
All parameters statistically significant. Can run acf to check adequacy. Also could run normality test, see if assumption is correct.

### b
Vel tas GARCH-M, nezinau kaip runinti

### c
```{r}
spec_3.8.3 <- ugarchspec(variance.model = list(model = "sGARCH", garchOrder = c(1,1)),
                         mean.model = list(armaOrder = c(0,0), include.mean = TRUE),
                         distribution.model = "std")
fit_3.8.3 <- ugarchfit(spec_3.8.3, log_data_3.8[,1])
fit_3.8.3
```

Checking hypothesis if degrees of freedom = 6. Calculating t statistic [9.4 (estimated degree of freedom) - 6 (checked degrees)] / 2.96 (std. error) = 1,15 < 1.96, therefore we can't reject null hypothesis that degrees of freedom = 6.

### d

```{r}
spec_3.8.4 <- ugarchspec(variance.model = list(model = "eGARCH", garchOrder = c(1, 1)),
                   mean.model = list(armaOrder = c(0, 0), include.mean = TRUE),
                   distribution.model = "norm")
fit_3.8.4 <- ugarchfit(spec_3.8.4, data = log_data_3.8[,1], solver = "hybrid")
fit_3.8.4
```
All parameters are statistically significant, there exists asymmetry.

### e
```{r}
fc_3.8.1 <- ugarchforecast(fit_3.8.1, n.ahead = 6)
fc_3.8.3 <- ugarchforecast(fit_3.8.3, n.ahead = 6)
fc_3.8.4 <- ugarchforecast(fit_3.8.4, n.ahead = 6)

FC_1 <- as.data.frame(fc_3.8.1@forecast$sigmaFor)
FC_3 <- as.data.frame(fc_3.8.3@forecast$sigmaFor)
FC_4 <- as.data.frame(fc_3.8.4@forecast$sigmaFor)

comparison <- FC_1 %>% bind_cols(FC_3) %>% bind_cols(FC_4)
names(comparison) <- c("GARCH-norm", "GARCH-stjud","EGARCH")
comparison
```
EGARCH decays faster.

# 6
a dalyje, tikrinam acf, how to remove autocorrelation - add MA part to mean equation of returns

# 10
checking summer effect. Prideti dummy variable prie paprasto GARCH, 1 - summer, 0 - ne.
checking GM effect - prideti kaip papildoma regresoriu i GARCH (bet t-1).

# 14
### a
Ar cia reikia ideti tik 3 ir 5 lagus i modeli? Per "fixed" kazkaip? 
```{r}
#AR(5)-GARCH(1,1)
data_3.14 <- read.table("d-gmsp9908.txt",header=T) 
Data_3.14 <- data_3.14 %>% 
  mutate(date = paste0(substr(date,1,4),"-",substr(date,5,6),"-",substr(date,7,8)))
log_data_3.14 <- log(Data_3.8[,2:3]+1)*100 # log returns

spec_3.14.1 <- ugarchspec(variance.model = list(model = "sGARCH", garchOrder = c(1, 1)),
                   mean.model = list(armaOrder = c(5, 0), include.mean = TRUE),
                   distribution.model = "ged")
spvol <- ugarchfit(spec_3.14.1, data = log_data_3.14[,2])
```
Gaunam AR(5) ir GARCH(1,1) forma S&P indeksui. Toliau b dalyje, sia sigma pridesim kaip exogenin i kintamaji i GARCH modeli GM grazoms.

### b 
ARMA(0,0) GARCH(1,1) + exogeninis (sigma S&P indekso)

```{r}
#reikia issiaiskinti kaip prideti external regressor i ugarchspec variance model! turetu buti df, reiktu gauti sigma values

# pavyzdys https://stackoverflow.com/questions/54293924/rugarch-external-regressors-in-mean-variance
spec_3.14.2 <- ugarchspec(variance.model = list(model = "sGARCH", garchOrder = c(1, 1),
                                                external.regressors = spvol),
                   mean.model = list(armaOrder = c(5, 0), include.mean = TRUE),
                   distribution.model = "ged")
spvol_2 <- ugarchfit(spec_3.14.12, data = log_data_3.14[,2])

#check adequacy, residualsu ljung box, acf/pacf
#stebim ar koeficientas bus reiksmingas, turbut, kad ne 

````
The inclusion of S&P volatility serves as a proxy for the market volatility. However, thismarket volatility does not contribute signﬁcantly to the modeling of GM stock returns.

# 15
Labai panasi eiga. Isvada:
No, the estimate of the parameter of ˆztis not signiﬁcantly diﬀerent from zero, with at-value of 0.083. Hence, the volatility of the GM returns does not have explanatorypower in the market volatility (in this case the S&P volatility.)