---
title: "Tsay_Uni_10_skyrius"
output: html_document
date: "2023-10-11"

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(fBasics)
library(markdown)
library(dplyr)
library(rugarch)
library(fGarch)
library(tseries)
library(MTS)
library(vars)
library(ggplot2)
library(urca)
library(tsDyn)
library(zoo)
library(rmgarch)
# library(ccgarch2, lib = '/private/var/folders/_p/tm4_6__j08j511mb0zbmlbg40000gn/T/Rtmp9DllBh/downloaded_packages')

```

# 1
```{r}
data_10.1<- read.table("m-ibmhpqsp6208.txt",header=T) 
# names(data_3.7) <- c("date", "3M")
Data_10.1 <- data_10.1 %>% 
  mutate(date = paste0(substr(date,1,4),"-",substr(date,5,6),"-",substr(date,7,8)))

plot(ts(Data_10.1)[,2:4])

# Define a multivariate GARCH model specification
 help(EWMAvol)
 ewma <- EWMAvol(Data_10.1[,2:4], lambda = -1)
plot(ts(ewma$Sigma.t))
```
Outcome: with EWMA volatility is smoother. Estimated lambda is 0.9672

# 2

```{r}
# var_model <- VAR(Data_10.1[,2:3], p = 1)

# Estimate a diagonal VEC model with lag order 1
diagonal_vec_model <- VECM(Data_10.1[,2:3], lag = 1)
summary(diagonal_vec_model)

#fitted volatility (ne tas...)
plot(ts(diagonal_vec_model$fitted.values), type = "line")


#Is the model adequate?
acf(diagonal_vec_model$residuals)
#Ljungbox

Box.test(ts(diagonal_vec_model$residuals)[,1], lag = 1, type = "Ljung")
Box.test(ts(diagonal_vec_model$residuals)[,1]^2, lag = 1, type = "Ljung")

Box.test(ts(diagonal_vec_model$residuals)[,2], lag = 1, type = "Ljung")
Box.test(ts(diagonal_vec_model$residuals)[,2]^2, lag = 1, type = "Ljung")

#p-value > 0.05 therefore we can't reject null hypothesis saying that there is no serial correlation
#between residuals. Altho there is in squared values.

#Multivariate Q statistic
portmanteau_test <- mq(diagonal_vec_model$residuals, lag = 6)
portmanteau_test <- mq(diagonal_vec_model$residuals^2, lag = 6)
#multivariate statistic rejects VECM model

#plot residuals 
acf(diagonal_vec_model$residuals)
# there is autocorrelation in first few lags. Model has to be improved
```

# 3
```{r}
speclist <- multispec(c(ugarchspec(distribution = "norm"), ugarchspec(distribution = "norm")))
bekk_spec <- dccspec(uspec = speclist,
                     dccOrder = c(1, 1),
                     model = "BEKK",
                     distribution = "mvnorm")

bekk_model <- dccfit(bekk_spec, data = Data_10.1[,3:4])
#issiimti lower triangular matrix is parametru listinamu

#fitted volatility and time-varying correlations
plot(bekk_model) #conditional sigma and correlation
```

#4 
```{r}
speclist <- multispec(c(ugarchspec(distribution = "norm"), ugarchspec(distribution = "norm"), ugarchspec(distribution = "norm")))
ccc_spec <- dccspec(uspec = speclist,
                    dccOrder = c(1, 1), model = "CCC", distribution = "mvnorm")

ccc_model <- dccfit(ccc_spec, data = Data_10.1[,2:4])

#volatilities
plot(ts(ccc_model@model$sigma))

conditional_correlations <- rcor(ccc_model)
rowMeans(conditional_correlations)
#correlations
ccc_model@mfit$R #imti pirma. Simetrinis

ccc_model@mfit$R[[564]]

#fitted model pavyzdys grazus solutionuose

# forecastinimas
forecast_horizon <- 1  # Specify the forecast horizon (1 for 1 step ahead)
ccc_forecast <- dccforecast(ccc_model, n.ahead = forecast_horizon)

# Extract the forecasted conditional covariance matrix
forecasted_covariance_matrix <- ccc_forecast@mforecast$R
```

# 5
Toks pat kaip 4

# 6

```{r}
speclist <- multispec(c(ugarchspec(distribution = "norm"), ugarchspec(distribution = "norm"), ugarchspec(distribution = "norm")))
dcc_spec <- dccspec(uspec = speclist,
                        dccOrder = c(1, 1),
                        distribution = "mvnorm")

# In this case, set it to 'DCC' for Dynamic Conditional Correlation

# Estimate the Cholesky GARCH model
DCC_model <- dccfit(dcc_spec, data = Data_10.1[,2:4])

plot(ts(DCC_model@model$sigma))
# View the model summary
summary(DCC_model)

DCC_model@mfit$R

```

#7
Pavyzdys solutionuose.
Paaiskinimas knygoj 466 psl.
Yra sasiuviny uzrasu

forecastinimo dalis: issitraukiam paskutini sigma11, sigma22, a1, a2, isitatom i apibrezta GARCHa ir gaunam final matrica.


#8 
Panasiai kaip 7, tik skiriasi rho forma. Cia cholesky deomposition.

#9
```{r}
speclist <- multispec(c(ugarchspec(distribution = "norm"), ugarchspec(distribution = "norm")))
dcc_spec <- dccspec(uspec = speclist,
                        dccOrder = c(1, 1),
                        distribution = "mvnorm")

# In this case, set it to 'DCC' for Dynamic Conditional Correlation

# Estimate the Cholesky GARCH model
DCC_model <- dccfit(dcc_spec, data = Data_10.1[,3:4])

# Obtain the conditional covariance matrix from the model
cov_matrix <- rcov(DCC_model)

# Perform Cholesky decomposition
chol_decomp <- chol(cov_matrix)

# The diagonal of chol_decomp contains the standard deviations of the two series
conditional_std_devs <- sqrt(diag(chol_decomp))

# The lower triangle of chol_decomp contains the Cholesky factorization of the conditional correlation matrix
conditional_correlations <- chol_decomp / (conditional_std_devs %*% t(conditional_std_devs))

# Print the conditional standard deviations
print("Conditional Standard Deviations:")
print(conditional_std_devs)

# Print the conditional correlation matrix
print("Conditional Correlation Matrix:")
print(conditional_correlations)
```

# 9
toks pat kaip 8, tik 3 dimensions


# 10
## Univariate
Knygoj 481 psl
fit univariate GARCH for each time series
calculate r1, sigma1^2, r2, sigma2^2, rho forecastui (t.y. isistatant paskutines datos values)
q1 = r1 - 1.65*(sigma1^2)^(1/2)
q2 = ...

VaR1 = 1000000*q1/100
VaR = sqrt(VaR1^2 + VaR2^2)

## Constant Correlation Bivariate
viskas tas pats tik fittinam diagonal bivariate GARCH (kaip 10.2)
cia kitoks rho ir turim gauti forecastintas vertes is to modelio

## Time-varying correlation model
kaip 10.8, fittinam per cholesky decomposition. 
Gaunam forecastintas vertes r1, r2, g11, q21, g22.
Is ju gaunam sigma1, sigma21, sigma2 ir corr.coef.
Toliau skaiciuojam VaR kaip pirmai.