---
title: "Tsay_Uni_8_skyrius"
output: html_document
date: "2023-09-19"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(fBasics)
library(markdown)
library(dplyr)
library(rugarch)
library(fGarch)
library(tseries)
library(MTS)
library(vars)
library(ggplot2)
library(urca)
library(tsDyn)
```

# 1
### a

```{r}
data_8.1 <- read.table("m-mrk2vw.txt",header=T) 
# names(data_3.7) <- c("date", "3M")
Data_8.1 <- data_8.1 %>% 
  mutate(date = paste0(substr(date,1,4),"-",substr(date,5,6),"-",substr(date,7,8)))

# Calculate the sample mean of log returns
mean_returns <- apply(Data_8.1[, 2:7], MARGIN = 2,FUN = mean)
mean_returns

# st.dev
st_dev<- apply(Data_8.1[, 2:7], MARGIN = 2,FUN = sd)
st_dev

# Calculate the covariance matrix of log returns
cov_matrix <- cov(Data_8.1[, 2:7])
cov_matrix

# Calculate the correlation matrix of log returns
cor_matrix <- cor(Data_8.1[, 2:7])
cor_matrix

```

Prideti komentaru, kas ka reiskia, kokios izvalgos

### b
```{r}
# Perform the multivariate Ljung-Box test
portmanteau_test <- mq(Data_8.1[, 2:7], lag = 6)
```
Therefore, one rejects the null hyphthesis of no serial or cross-sectional correlations in the data

### c
Because of the signiï¬cance of serial correlations in Part(b), there is lead-lag relationship among the six return series.

# 2
bivariate autoregressive model
pradzioj M(i) statistica ir AIC/BIC budu surandam lag'a
runinam koda 

```{r}
var_model <- VAR(ts_data, p = 2)

# Summary of the VAR model
summary(var_model)

#VMA

# Create a lag operator polynomial
lag_polynomial <- VARrep(1, p = q, type = "VMA")

# Fit the VMA model
vma_model <- VAR(ts_data, type = "none", lag.l = lag_polynomial)

# Summary of the VMA model
summary(vma_model)
```
Palyginti sigmas, ar panasios.

Del strukturines formos. Fittinta ir sutvarkyta modeli keicia.

Palikti tik significant lagai:

```{r}
data <- matrix(data = rnorm(200), nrow = 100, ncol = 2)
ts_data <- ts(data, start = c(2000, 1), frequency = 12)
var_model <- VAR(ts_data, p = 3)  # Example VAR model with 3 lags

# Specify the equation and lags to keep
equation_to_modify <- 1  # Replace with the equation index you want to modify (e.g., the first equation)
lags_to_keep <- 1:2     # Replace with the specific lags you want to keep (e.g., lags 1 and 2)

# Set the coefficients of unwanted lags to zero
for (lag in 1:var_model$p) {
  if (lag %in% lags_to_keep) {
    next  # Keep the specified lags, skip the rest of the loop
  }
  var_model$varresult$A[, , equation_to_modify][, lag] <- 0
}

# Optionally, you can update the VAR model name to reflect the changes
var_model$name <- paste("VAR with specific lags for equation", equation_to_modify)

# Summary of the modified VAR model
summary(var_model)
```

strukturine forma:
choleskio dekompozicija identifyina strukturinius sokus:

```{r}
# Example: Estimating a structural VAR with Cholesky decomposition
# Assuming you have already estimated a VAR model named 'var_model'

# Perform Cholesky decomposition to identify structural shocks
svar_model <- SVAR(var_model, type = "Cholesky")

# Summary of the structural VAR model
summary(svar_model)

# Plot the impulse response functions
plot(svar_model)

```

# 4
```{r}
# Perform the threshold cointegration test
toda_test_result <- toda.test(ts1, ts2)

# Print the test results
print(toda_test_result)
```

if p<0.05 they are threshold cointegrated. We further build thershold model

```{r}
ts1 <- cumsum(rnorm(n))
ts2 <- cumsum(rnorm(n))
# ar cia tikrai cumsum reikia?

# Estimate the threshold bivariate model
threshold_model <- TVECM(cbind(ts1, ts2), lag = 1, threshold = "linearity")

# Print the model summary
print(summary(threshold_model))

# Plot the threshold effects
plot(threshold_model)
```



# 7

### a
```{r}
data_8.7 <- read.table("m-gs1n3-5304.txt",header=F) 
names(data_8.7) <- c("Y1", "Y3", "year", "month", "day")
Data_8.7 <- data_8.7 %>% 
  mutate(date = paste0(year,"-",month,"-",day),
         date = as.Date(date, format = "%Y-%m-%d")
         # Y1 = log(Y1),
         # Y3 = log(Y3)
         ) %>% 
  dplyr::select(Y1,Y3,date)

ts_Data_8.7 <- ts(Data_8.7[,1:2])
plot(ts_Data_8.7)

#In order to define lag order we use AIC/BIC
VARselect(ts_Data_8.7, lag.max = 6, type = "none")

var_1 <- VAR(ts_Data_8.7,p=4)
summary(var_1)

acf(residuals(var_1))

# residuals show that model could be further improved, but it seems to capture the dynamic struxture of the data reasonably well.

### sasiuvinyje turiu apsirasiusi formas. Isvada - unidirectional. 370 psl knygos yra grazus aprasymas


# Compute the impulse response functions (IRFs)
irf_results <- irf(var_1, impulse = "Y1", response = c("Y1", "Y3"), n.ahead = 6)

# Plot the IRFs
plot(irf_results)
```
Aprasyti ka rodo grafikas. Kiek lagu islieka soko itaka kuriam kintamajam ir kokiam lygyje.

### c
```{r}
forecast_horizon <- 12
forecast_results <- predict(var_1, n.ahead = forecast_horizon)

### reikia prideti 
# Create a data frame for the forecasted values and confidence intervals
forecast_data <- data.frame(
  Period = 1:forecast_horizon,
  # Actual = Data_8.7[,1], # Replace with your actual Y1 data
  Forecast = forecast_results$fcst$Y1[,1],
  Lower_CI = forecast_results$fcst$Y1[,2],
  Upper_CI = forecast_results$fcst$Y1[,3]
)

# Create the ggplot2 plot
ggplot(forecast_data, aes(x = Period)) +
  # geom_line(aes(y = Actual), color = "black", linetype = "solid") +
  geom_line(aes(y = Forecast), color = "red", linetype = "dashed") +
  geom_ribbon(aes(ymin = Lower_CI, ymax = Upper_CI), fill = "blue", alpha = 0.3) +
  labs(
    title = "Forecasted Values for Y1",
    y = "Y1",
    x = "Period"
  ) +
  theme_minimal()

#reiketu prideti grafika
```

### d
#augmented dickey fuller test irgi variantas
```{r}
# Specify the presence of restricted constant terms
const_vec <- c("const", "const")

# Perform Johansen cointegration test with restricted constant terms (no drift in the mean)
#ideja, tikrinti rank(Pi)
johansen_test_result <- ca.jo(data.frame(ts_Data_8.7[,1], ts_Data_8.7[,2]), type = "eigen", ecdet = "const", K = 2)


# Print test results
summary(johansen_test_result)
plot(johansen_test_result)
```



### e
```{r}
vecm <- tsDyn::VECM(ts_Data_8.7, lag=2)
summary(vecm)
```

Kaip surasyti matematine forma (sasiuvinyje)

The ECT matrix and its coefficients indicate how each variable adjusts back to its long-run equilibrium when deviations occur. If beta1  and beta2  are both statistically significant and have opposite signs, it suggests cointegration between the variables, and the ECT represents the speed of adjustment toward the equilibrium relationship.

Error Correction Term (ECT): The ECT represents the speed at which the system corrects deviations from the long-term equilibrium. A significant ECT coefficient indicates the strength of this adjustment. For instance, if the ECT coefficient for GDP in the consumption equation is -0.2, it suggests that in the short term, if consumption deviates from the long-term equilibrium by 1%, it will adjust by approximately 0.2% towards the equilibrium in the following period.

### f
```{r}
forecast_results_vecm <- predict(vecm, n.ahead = forecast_horizon)
#reiketu dar grafiko
```

### g
```{r}
forecast_results
forecast_results_vecm



#ideti bendrai palyginti grafike. kuris duoda didesni forecasta ir kuris mazesni standard error?
```
